Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-09 WholePlotTrajectories.tex
relecture

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity, but this intermediate disturbance hypothesis (IDH) remains
debated, and this controversy also questions the extent of communities
resilience regarding their functional and taxonomic facets. To
disentangle the ecological processes driving community response to
disturbance, we analysed the taxonomic and functional diversity
trajectories following a disturbance gradient. Specifically, we
examined, over 30 years, the functional and taxonomic community
trajectories with regards to diversity, composition and redundancy.
Functional trajectories were drawn based on 7 leaf, stem and
life-history traits. We highlighted the cyclic recovery of community
taxonomic and functional composition. The pre-disturbance taxonomic
differences were maintained over time while the functional composition
trajectories were quite similar. The IDH did predict communities
functional diversity response while taxonomic diversity remained poorly
sensitive to disturbance intensity. Although consistent, the recovery of
community composition, diversity and redundancy remained unachieved
after 30 years. This acknowledged the need of decades-long cycles with
no disturbance to ensure a complete recovery, and questioned tropical
forest community resilience after repeated disturbance.
}

@@

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. Specifically, we (i) questioned the
recovery of communities taxonomic and functional characteristics and
identified the underlying assembly processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and elucidated
its translation into different trajectories, and (iii) questioned
community recovery time.

@@

From 1989 (2 years after disturbance) to 2015 (28 years after
disturbance), 828 388 individual trees and 591 botanical species
pertaining to 223 genus and 64 families were recorded. For undisturbed
plots, taxonomic Richness and Evenness remained stable over the 30 years
monitored. In disturbed communities, after low disturbance intensity the
taxonomic richness increased, reaching a maximum gain of 14 botanical
genera (plot 3 from treatment 2). After intense disturbance the
taxonomic richness followed a more complex trajectory, decreasing for
ten years after disturbance before recovering to pre-disturbance values.
The maximum richness loss or gain after disturbance was positively
correlated to the disturbance intensity
(\(\rho_{spearman}^{Richness}=0.50\)). In all disturbed plots the
taxonomic evenness first increased until a maximum reached after around
20 years. This maximum was positively correlated to the disturbance
intensity (\(\rho_{spearman}^{Evenness}=0.77\)). The evenness then
stabilized except for two T3 plots (plots 8 and 12) for which evenness
kept increasing.

@@

@@

@@

Conversely, disturbed communities followed functional trajecotories
similar in terms of functional composition. As pre-disturbance surviving
trees mirror the initial community \citep{Herault2018}, changes in
functional composition relied upon the recruitment of species or
functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches
of high availability of light, space and nutrient vacated by the
disturbance. The recruitment of pioneers similarly changed community
functional composition towards more resource-acquisitive strategies
\citep{Grime1998, Westoby1998, Wright2004, Reich2014}. Thereafter, again
in the same way for all disturbed communities, long-lived, more
resistant and shade-tolerant species excluded the first established
pioneers and started the recovery of pre-disturbance functional
composition.

@@

The IDH predicted well the disturbance impact on community taxonomic
richness, enhanced until an intensity threshold (20-25\% AGB loss), and
to some extent on taxonomic evenness, somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.). Below the intensity
threshold the size of the surviving community maintained the
pre-disturbance taxonomic richness while the recruitment of pioneers,
infrequent or absent before disturbance, increased both community
taxonomic richness and evenness. Beyond the intensity threshold, the
disturbance decreased the taxonomic richness of surviving trees which
was not offset by the enrichment of pioneers, so that the overall
community taxonomic richness decreased according to the disturbance
intensity. For community taxonomic evenness the disturbance impact was
similar but slighter, as the evenness is less sensitive to the loss of
rare species. Taxonomic evenness rather represented the increasing
dominance of pionneers that balanced the usual hyper-dominance of
tropical forest below the intensity threshold, thus increasing community
overall evenness up to the intensity threshold beyond which pioneers
became in turn highly dominant and decreased the overall evenness.

Conversely the IDH was disproved regarding the disturbance impact on
community functional diversity. Irrespective of the disturbance
intensity the recruitment of pionneers, functionally highly different
from the composition of pre-disturbance community, increased the
community functional richness and eveness.

Along time, taxonomic richness trajectories for all disturbed
communities first dropped similarly following the loss of species during
disturbance. Then, trajectories showed a species gain depending on the
disturbance intensity. The species gain was all the more significant as
the disturbance intensity increased, up to an intensity threshold,
through the establishment of long-lived pioneers enhancing community
taxonomic richness and evenness in the long term. These long-lived
pionneers, functionally quite different from the functional composition,
entailed in parallel a progressive and long-lasting increase of the
functional diversity. Beyond an intensity threshold, though, a few
short-lived pioneers occupied the vacated environmental space and
prevented the establishment of other species. These short-lived pioneers
were functionally very different from the pre-disturbance community and
entailed a rapid and significant increase of functional richness en
evenness. Already after 10 years, though, short-lived pioneers started
to decline and the functional richness and evenness decreased. Likely
the establishment of long-lasting pioneers will follow this decrease and
we expect the taxonomic and functional trajectories to catch up with
those observed after intermediate disturbance.

@@

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. The redundancy
decrease was not compensated in the first place as the pionneers first
recruited were functionally different from the pre-disturbance
functional composition. Progressively though, first established species
were replaced by more competitive long-lived pionneers or
late-successional species resembling more the pre-disturbance functional
composition and restoring the functional redundancy. This replacement
was stochastic and followed the lottery recruitment rules, implying a
recruitment eased for the first recruited species but then increasingly
hampered by the emergence of interspecific competition
\citep{Busing2002}. Along time the recovery of infrequent species was
increasingly difficult, so that the time for the full recovery of the
functional redundancy, in some communities just initiated after 30
years, was difficult to estimate
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

@@

Our study revealed the post-disturbance recovery of both taxonomic
composition, maintaining the initial differences among communities, and
functional composition, that was in contrast similar among communities.
The IDH was validated for community taxonomic richness and, to some
extent, taxonomic evenness but disproved regarding community functional
richness and evennes that were enhanced for any disturbance intensity by
the high functional differences of pioneers compared to
late-successional functoinal composition. The IDH was translated in time
by the recruitment, beyond an intensity threshold, of short-lived
pionneers that prevented in the first times the establishment of more
diverse long-lived pionneers, recruited otherwise below the intensity
threshold. The resilience of tropical forests proved consistent although
several decades-long. Still, the disturbance impact on communities
redundancy cautioned against the risks of infrequent species loss and
the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}. As the trajectories highlighted the
recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.

>>>>

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH),though,
remains debated and the controversy questions the extent of communities'
resilience regarding their functional and taxonomic facets. To
disentangle the ecological processes driving community response to
disturbance, we analysed the diversity trajectories over 30 years
following a disturbance gradient. Specifically, we examined community
functional and taxonomic trajectories with regards to diversity,
composition and redundancy. Functional trajectories were drawn based on
7 leaf, stem and life-history traits. We highlighted the cyclic recovery
of community taxonomic and functional composition. While pre-disturbance
taxonomic differences were maintained over time, functional composition
trajectories were quite similar among communities. The IDH did predict
communities taxonomic diversity response while functional diversity was
enhanced whenever the disturbance intensity. Although consistent, the
recovery of community composition, diversity and redundancy remained
unachieved after 30 years. This acknowledged the need of decades-long
cycles with no disturbance to ensure a complete recovery, and questioned
tropical forest community resilience after repeated disturbance.
}

@@

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. Specifically, we (i) elucidated
community taxonomic and functional recovery and the determinism of
underlying processes, (ii) clarified the validity of the IDH in the long
term for tropical forest and its translation into different trajectories
in time, and (iii) questioned community recovery time.

@@

@@

From 1989 (2 years after disturbance) to 2015 (28 years after
disturbance), 828 388 individual trees and 591 botanical species
pertaining to 223 genus and 64 families were recorded.

@@

For undisturbed plots, taxonomic Richness and Evenness remained stable
over the 30 years monitored. In disturbed communities, after low
disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2). After
intense disturbance the taxonomic richness followed a more complex
trajectory, decreasing for ten years after disturbance before recovering
to pre-disturbance values. The maximum richness loss or gain after
disturbance was positively correlated to the disturbance intensity
(\(\rho_{spearman}^{Richness}=0.50\)). In all disturbed plots the
taxonomic evenness first increased until a maximum reached after around
20 years. This maximum was positively correlated to the disturbance
intensity (\(\rho_{spearman}^{Evenness}=0.77\)). The evenness then
stabilized except for two T3 plots (plots 8 and 12) for which evenness
kept increasing.

@@

Conversely, disturbed communities followed functional trajectories
similar in terms of functional composition. As pre-disturbance surviving
trees mirror the initial community \citep{Herault2018}, changes in
functional composition relied upon the recruitment of species or
functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches
of high availability of light, space and nutrient vacated by the
disturbance. The recruitment of pioneers changed community functional
composition in the same way for all disturbance intensity towards more
resource-acquisitive strategies, moving community functional composition
right along the first axis in Figure \ref{fig:NMDSplans}
\citep{Grime1998, Westoby1998, Wright2004, Reich2014}. Thereafter
long-lived, more resistant and shade-tolerant species excluded the first
established pioneers and started the recovery of pre-disturbance
functional composition, moving similarly community functional
composition left along the first axis and upward along the second axis
in Figure \ref{fig:NMDSplans} .

These trajectories provided empirical support to the hypothesis that
community assembly is both deterministic and historically convergent at
different levels of community organization. Deterministic, trait-based
processes drove community convergence in functional composition, while
at the same time dispersal limitation maintained their divergence in
taxonomic composition \citep{Fukami2005}.

@@

The IDH predicted well the disturbance impact on community taxonomic
richness, enhanced until an intensity threshold (20-25\% AGB loss), and
to some extent on taxonomic evenness, somewhat decoupled from the
disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.)
\citep{Zalamea2012, Guitet2018}. Below the intensity threshold the size
of the surviving community maintained the pre-disturbance high taxonomic
richness while the recruitment of pioneers, infrequent or absent before
disturbance, increased both community taxonomic richness and evenness.
Beyond the intensity threshold, the disturbance decreased the taxonomic
richness of surviving trees which was not offset by the enrichment of
pioneers, so that the overall community taxonomic richness decreased
according to the disturbance intensity \citep{Molino2001}. For community
taxonomic evenness the disturbance impact was similar but slighter, as
the evenness is less sensitive to the loss of rare species. Taxonomic
evenness rather represented the increasing dominance of pionneers that
balanced the usual hyper-dominance of tropical forest below the
intensity threshold, thus increasing community overall evenness up to
the intensity threshold beyond which pioneers became in turn highly
dominant and decreased the overall evenness \citep{Baraloto2012a}.

Conversely the IDH was disproved regarding the disturbance impact on
community functional richness and evenness. Irrespective of the
disturbance intensity the recruitment of pionneers, functionally highly
different from the composition of pre-disturbance community, increased
both community functional richness and eveness.

Along time, taxonomic richness trajectories of all disturbed communities
first dropped similarly, following the species loss due to disturbance,
and then displayed a species gain depending on the disturbance
intensity. Up to an intensity threshold, the species gain was all the
more significant that the disturbance intensity increased, with the
establishment of long-lived pioneers enhancing community taxonomic
richness and evenness in the long term. These long-lived pionneers,
functionally quite different from the functional composition, entailed
as well a progressive and long-lasting increase of the functional
richness and evenness \citep{Denslow1980, Molino2001}. Beyond an
intensity threshold, though, a few short-lived pioneers occupied the
vacated environmental space and prevented the establishment of other
species. These short-lived pioneers were functionally very different
from the pre-disturbance community and entailed a rapid and significant
increase of functional richness en evenness. Already after 10 years,
though, short-lived pioneers started to decline and the functional
richness and evenness decreased. Likely this decrease will be followed
by the establishment of long-lasting pioneers, and by the time they
recruit we expect the taxonomic and functional trajectories to catch up
with those observed after intermediate disturbance
\citep{Anderson2007, Walker2009}.

@@

For 15 years the species loss during disturbance, determined by the
disturbance intensity, commensurately decreased the functional
redundancy within the pre-disturbance functional space. The redundancy
decrease was not compensated in the first place as the first recruited
pionneers were functionally different from the pre-disturbance
functional composition. Progressively though, first established species
were replaced by more competitive long-lived pionneers or
late-successional species resembling more the pre-disturbance functional
composition and restoring the functional redundancy. This replacement
was stochastic and followed the lottery recruitment rules, implying a
recruitment eased for the first recruited species but then increasingly
hampered by the emergence of interspecific competition
\citep{Busing2002}. Along time the recovery of infrequent species was
increasingly difficult, so that the time for the full recovery of the
functional redundancy, in some communities just initiated after 30
years, was difficult to estimate
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

@@

Our study revealed community recovery through the combination of
deterministic processes driving their convergence in functional
composition, and dispersal limitation maintaining their divergence in
taxonomic composition. The IDH was validated for community taxonomic
richness and, to some extent, taxonomic evenness but disproved regarding
community functional richness and evennes that were enhanced for any
disturbance intensity by the high functional differences of pioneers
compared to late-successional functoinal composition. The IDH was
translated in time by the recruitment, beyond an intensity threshold, of
short-lived pionneers that prevented in the first times the
establishment of more diverse long-lived pionneers, recruited otherwise
below the intensity threshold. The resilience of tropical forests proved
consistent although several decades-long. Still, the disturbance impact
on communities redundancy cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Gourlet-Fleury2005}. As the trajectories highlighted the
recruitment processes proved central for communities response to
disturbance and closer focus on demographical drivers of communities
response would clarify the fate of the future forests.
